---
title: "Tidy Tuesday 2026, Week 5: Edible Plants Database"
author: "Author: Lyndsay Miles"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data source for this week's visualizations: <https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-02-03/readme.md>

#tidytuesday \| 2026 \| Wk5 \| Edible Plants Database

This week our challenge was to edible plants courtesy of the GROW Observatory's Edible Plant Database (EPD). The GROW Observatory is a European Citizen Science project regarding growing food, soil moisture sensing and land monitoring. Thanks to Nicola Rennie, who curated this week's dataset.

For this challenge, I decided to look at edible plants in terms of 3 common considerations (temperature tolerance, water required, and seasonality) through a Sankey chart using the ggalluvial() R package. Since Sankey charts cannot tell us much about a particular plant, I also created an interactive chart using reactable().


```{r, echo=TRUE, warning=FALSE, message=FALSE, include = FALSE}
edible_plants <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-02-03/edible_plants.csv')
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(tidyverse)
library(ggthemes)
library(ggalluvial)
library(reactable)
library(htmltools)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, include = FALSE}
#note: data cleaning not shown
df <- edible_plants %>% 
  mutate(water_cat = case_when(water == "very high" ~ "Very High",
                               water == "Medium" ~ "Medium",
                               water == "Very low" ~ "Very Low",
                               TRUE ~ water
                               ),
         season_cat = case_when(season %in% 
                                  c("biennial","biennial, grown as annual", "Biennial, grown as an annual") ~ "Biennial",
                                season %in% c("Annual", "Annual/perannial") ~ "Annual",
                                season %in% c("Perennial", "perennial", "Perrenial", "Perrenial evergreen", "Semi-evergreen perrenial") ~ "Perennial",
                                TRUE ~ NA),
         temperature_class = case_when(
           temperature_class == "Very hard" ~ "Very hardy", 
           TRUE ~ temperature_class)
  ) %>%
  mutate(water_cat = factor(water_cat,
                          levels = c("Very Low", "Low", "Medium", "High", "Very High")),
         temperature_class = factor(temperature_class,
                                    levels = c("Very hardy", "Half hardy", "Hardy",
                                               "Tender", "Very tender"))
         )
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, include = FALSE}
flow_data <- df %>%
  filter(season_cat != "unknown") %>%
  count(temperature_class, water_cat, season_cat, name = "n")
```

```{r, echo = TRUE, warning=FALSE, message=FALSE}
#data cleaning not shown
ggplot(flow_data,
       aes(axis1 = temperature_class,
           axis2 = water_cat,
           axis3 = season_cat,
           y = n)) +

  geom_alluvium(aes(fill = temperature_class), width = 1/12, alpha = 0.8) +
  geom_stratum(width = 0.25, fill = "#E9E3D6", color = "#B8B0A3")+
  geom_text(stat="stratum",
          aes(label = after_stat(stratum)),
          vjust = 0.7,
          size =3.25)+
  scale_x_discrete(limits = c("Temperature", "Water Required", "Season"),
                   expand = c(.05, .05)) +
  labs(
    title = "What Edible Plants Need to Thrive",
    subtitle = "From Temperature Tolerance to Water and Season",
    y = "Number of plants",
    x = NULL,
    fill = "Temperature",
    caption = "Lyndsay Miles | #TidyTuesday\nSource: the GROW Observatory"
    ) +
  scale_fill_manual(values = c(
  "Very hardy" = "#A7C957",
  "Hardy"      = "#2A9D8F",
  "Half hardy" = "#264653",
  "Tender"     = "#E76F51"
))+

  theme_solarized()+
  theme(
  plot.title = element_text(size = 16, face = "bold", color = "#3A3A3A"),
  plot.caption = element_text(size = 8, color = "#555555"),
  plot.subtitle = element_text(color = "#555555"),
  legend.title = element_text(color = "#555555"),
  legend.text  = element_text(color = "#555555")
)
```


```{r, echo=TRUE, warning=FALSE, message=FALSE, include=FALSE}
table <- df %>% select(
  -c(water, season) 
) %>%
  rename(season = season_cat,
         water = water_cat) %>% 
  mutate(plant_name = str_c(common_name, " (", taxonomic_name, ")"),
         preferred_ph_range = str_c("(", preferred_ph_lower, " , ", preferred_ph_upper, ")")) %>%
 select(
  plant_name, season, water, sunlight, temperature_class,
   soil, nutrients, preferred_ph_range, sensitivities
  )
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
#note: data cleaning not shown
tagList(
  tags$h3("Plant growing guide", style = "color:#586E75;"),
  tags$p("Water, sunlight, soil, and hardiness requirements", style = "color:#777777; margin-top:-6px;"),
  
reactable(
  table,
  filterable = TRUE,
  searchable = TRUE,
  pagination = TRUE,
  compact = TRUE,
  striped = TRUE,

  columns = list(
    plant_name = colDef(name = "Plant name"),
    season = colDef(name = "Season"),
    water = colDef(name = "Water"),
    sunlight = colDef(name = "Sunlight"),
    temperature_class = colDef(name = "Hardiness"),
    soil = colDef(name = "Soil"),
    nutrients = colDef(name = "Nutrients needed"),
    preferred_ph_range = colDef(name = "Ph Range"),
    sensitivities = colDef(name = "Sensitivities")
    
  ),

  defaultColDef = colDef(
    headerStyle = list(fontWeight = "600")
  ),
  style = list(background = "#FDF6E3")
)
)

```

